---
title: Class 2 - data types, daa loading, data manipulation
author: Szymon Drobniak
date: '`r Sys.Date()`'
format:
    html:
        toc: true
        toc-location: left
        toc-depth: 2
        theme: simplex
        embed-resources: true
        code-fold: true
        code-tools: true
        number-sections: true
        code-block-bg: "#f4f4f4"
        code-block-border-left: false
    gfm: default
crossref:
    fig-title: Figure
    tbl-title: Table
    title-delim: —
    fig-prefix: Fig.
    tbl-prefix: Tab.
    eq-prefix: Eq.
editor_options:
    chunk_output_type: console
execute:
    cache: true
    warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(here)
```

**In the below instructions...**

**EXERCISE 1:** are bits of code to execute/practice pieces to do, often with only hints on how to perform them.

**Output** indicates the typical output you should expect from a given instruction.

## Data

We will work on a simple dataset on cholesterol levels from patients. The data presents cholesterol concentrations in plasma in patients before diet, and after 4 & 8 weeks of diet containing one of two types of margarine. The age group of patients is also indicated. Data can be downloaded here: https://raw.githubusercontent.com/wbabik/Practical_computing/teaching/Class_10/data/Cholesterol_Age_R.csv

```{r echo = T}
data_chol <- read.table('https://raw.githubusercontent.com/wbabik/Practical_computing/teaching/Class_10/data/Cholesterol_Age_R.csv',
                       sep = ';', header = T,
                       stringsAsFactors = T)

head(data_chol)
summary(data_chol)
```
