---
title: Class 2 - data types, daa loading, data manipulation
author: Szymon Drobniak
date: '`r Sys.Date()`'
format:
    html:
        toc: true
        toc-location: left
        toc-depth: 2
        theme: simplex
        embed-resources: true
        code-fold: true
        code-tools: true
        number-sections: true
        code-block-bg: "#f4f4f4"
        code-block-border-left: false
    gfm: default
crossref:
    fig-title: Figure
    tbl-title: Table
    title-delim: —
    fig-prefix: Fig.
    tbl-prefix: Tab.
    eq-prefix: Eq.
editor_options:
    chunk_output_type: console
execute:
    cache: true
    warning: false
    echo: false
---

```{r setup, include=FALSE}
library(here)
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
```

::: {.callout-tip}
## Before you start...

Make sure you have the following set and ready:

- Folder you want touse as your working directory exists on your HDD
- Set this folder as working directory (using a menu or `setwd("YOUR_PATH)`)
- Relevant data (if provided) is in your working directory
:::

**In the below instructions...**

**EXERCISE 1:** are bits of code to execute/practice pieces to do, often with only hints on how to perform them.

**Output** indicates the typical output you should expect from a given instruction.



## Paths in R

Paths in R can be relative or absolute. For example, in my environment my current working directory is 

```{r}
#| echo: !expr format %in% c("html")


getwd()
```

Which is very long and hard to rememeber/type in. However, once set I can refer to files in this folder using *relative* paths (skipping the part leading TO the root directory). E.g. for the `class2` folder I can just write `"class2"` or `"./class2"`.
